The Unicode Cookbook \linebreak for Linguists

### 2015-11-15 book/chapters/introduction.tex
minor changes

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
actually an $8+4=12$-bit encoding, though with very many duplicates by design,
namely all ASCII codes were repeated in each 8-bit code page. To be precise,
ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible
characters. However, because of overlap and not-assigned codes points the actual
number of symbols was much smaller.} This system almost immediately was
understood to be insufficient and impractical, so various initiatives to extend
and reorganize the encoding started in the 1980s. This led, for example, to
various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
(e.g.~Mac OS Roman), which one still sometimes encounters today.

More wide-ranging, various people in the 1980s started to develop true
international code sets. In the United States, a group of computer scientists
formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991
(with $2^{16} = 65,536$ different possible characters). At the same time in
Europe, the \textsc{International Organization for Standardization} (ISO) was
working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with
theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode
Consortium and the ISO jointly develop the \textsc{unicode standard}, or
ISO/IEC~10646, leading to the current system including the original 16-bit
Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional
planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} =
1,114,112$ possible characters). The most recent version of the Unicode Standard
(currently at version number 7.0) was published in June 2014 and it defines
112,956 different characters \citep{Unicode2014}.

@@

\noindent In the next section we provide a very brief overview of the linguistic terminology
concerning writing systems before turning to the slightly different
computational terminology in the next section on the Unicode Standard.

>>>>

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
actually an $8+4=12$-bit encoding, though with very many duplicates by design,
namely all ASCII codes were repeated in each 8-bit code page. To be precise,
ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
characters. However, because of overlap and not-assigned codes points the actual
number of symbols was much smaller.} This system almost immediately was
understood to be insufficient and impractical, so various initiatives to extend
and reorganize the encoding started in the 1980s. This led, for example, to
various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
(e.g.~Mac OS Roman), which one still sometimes encounters today.

More wide-ranging, various people in the 1980s started to develop true
international code sets. In the United States, a group of computer scientists
formed the \textsc{unicode consortium}, proposing a 16-bit encoding in 1991
(with $2^{16} = 65,536$ different possible characters). At the same time in
Europe, the \textsc{international organization for standardization} (ISO) was
working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of
the \textsc{universal character set} (UCS) in 1990 was 31-bit (with
theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some
technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode
Consortium and the ISO jointly develop the \textsc{unicode standard}, or
ISO/IEC~10646, leading to the current system including the original 16-bit
Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional
planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} =
1,114,112$ possible characters). The most recent version of the Unicode Standard
(currently at version number 7.0) was published in June 2014 and it defines
112,956 different characters \citep{Unicode2014}.

@@

\noindent In the next section we provide a very brief overview of the linguistic
terminology concerning writing systems before turning to the slightly different
computational terminology in the subsequent section on the Unicode Standard.

### 2015-11-15 book/chapters/ipa_background.tex
minor changes

Established in 1886, the \textit{International Phonetic Association} (henceforth
\textit{Association}) has long maintained a standard alphabet, \textit{the
International Phonetic Alphabet} (henceforth \textit{IPA}), which is a common
standard in linguistics to transcribe sounds of spoken languages. It was first
published in 1888 as an international system of phonetic transcription for oral
languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

@@

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textit{Handbook of the International Phonetic
Association: A guide to the use of the International Phonetic Alphabet}
\citep{IPA1999}, henceforth simply called \textit{Handbook}.

@@

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (Chao tones). Importantly, however, this was
the first revision to address issues of computational representations for the
IPA -- the principles above of which have had several ramifications that make
the interoperability of electronic linguistic data extremely difficult.

>>>>

Established in 1886, the \textsc{international phonetic association} (henceforth
\textit{Association}) has long maintained a standard alphabet, the
\texsct{international phonetic alphabet} (henceforth \textit{IPA}), which is a
common standard in linguistics to transcribe sounds of spoken languages. It was
first published in 1888 as an international system of phonetic transcription for
oral languages and for pedagogical purposes. It contained phonetic values for
English, French and German. Diacritics for length and nasalization were already
present in this first version, and the same symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

@@

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems. The current construction and use of the IPA are guided by
principles outlined in the \textsc{handbook of the international phonetic
association} \citep{IPA1999}, henceforth simply called \textit{Handbook}.

@@

In 1989 an IPA revision convention was held in Kiel, Germany. As per other
revisions, there was expansion and revisions to phonetic symbols in the IPA
chart. Notably the marking of tone was extended with the addition of a second
system for marking linguistic tones (the so-called `Chao' tone letters).
Importantly, however, this was also the first revision to address issues of
computational representations for the IPA -- the principles above of which have
had several ramifications that make the interoperability of electronic
linguistic data extremely difficult.
